# Logistic regression

Load libraries

```{r}

library(tidyr)
library(dplyr)
library(ggplot2)

```

1. Read the data from local folder and check for structure and dimensions. Print out the names of the variables in the data and describe the data set briefly, assuming the reader has no previous knowledge of it. 

```{r}
alc <- read.table("C:/LocalData/kpuukko/IODS-project/data/alc.txt", sep="\t")

#There are 382 observations and 35 variables in the data set
str(alc)
dim(alc)
summary(alc)

#Names of the variables
colnames(alc)

table(alc$sex) #198 Female and 184 Male students
table(alc$age) # Age distribution is from 15 to 22, most of the students are 16 or 17 years old. 
table(alc$high_use) #114 of students are high alcohol users, 268 are not

# A bar plot of each variable
gather(alc) %>% ggplot(aes(value)) + facet_wrap("key", scales = "free") + geom_bar()

```
2. I will study the relationship between gender, school absenses, going out, famrel and alcohol consumption

Hypothesis

H1: High quality of family relationships reduce the risk for high alcohol consumption. 
H2: School absenses increase the risk for high alcohol consumption. 
H3: Going out increases the risk for high alcohol consumption.
H4: Male students use more alcohol than female students.  

3. Numerically and graphically explore the distributions of your chosen variables and their relationships with alcohol consumption (use for example cross-tabulations, bar plots and box plots). Comment on your findings and compare the results of your exploration to your previously stated hypotheses. (0-5 points)

```{r}
# A box plot of high_use and family relationship
#As assumed, students who are high alcohol users have lower quality family relationships. 
g1 <- ggplot(alc, aes(x = high_use, y = famrel))
g1 + geom_boxplot() + ylab("Family relationship")

# A box plot of high_use and school absenses
#Students who consume more alcohol have slightly more school absenses compared to students who are not high users. However, the differences between the groups appears really small. 
g2 <- ggplot(alc, aes(x = high_use, y = absences))
g2 + geom_boxplot() + ylab("School absenses")

# A box plot of high_use and going out
#As assumed, students who are high alcohol users go out more than other students.  
g3 <- ggplot(alc, aes(x = high_use, y = goout))
g3 + geom_boxplot() + ylab("Going out")

# A box plot of alcohol use and going out
#As assumed, men use alcohol more than women.  
g4 <- ggplot(alc, aes(x = sex, y = alc_use))
g4 + geom_boxplot() + ylab("Alcohol use")
```
4. Logistic regression analysis with family relationships, absences, going out and gender as predictors and high alcohol consumption as an outcome.  Present and interpret a summary of the fitted model. Present and interpret the coefficients of the model as odds ratios and provide confidence intervals for them. 

```{r}

#Logistic regression analysis with family relationships, absences, going out and gender as predictors and high alcohol consumption as an outcome. 
##The model confirmed that all the above variables are all significant predictors of high alcohol consumption 

Mod1 <- glm(high_use ~ famrel + absences + goout + sex, data = alc, family = "binomial")
summary(Mod1)

#print out coefficients
coef(Mod1)

#compute odds ratio
OR <- coef(Mod1) %>% exp

#computer confidence intervals 
CI <- confint(Mod1) %>% exp

## print out the odds ratios with their confidence intervals
cbind(OR, CI)

#The odds ratios of school absences, going out and sex male are above 1, meaning that they are positively related with high alcohol consumption.
#Interestingly the odds ratio of family relationship was below 1. If i understood right, this means that there is lesser likelihood that family relationships are associated with high alcohol consumption.
```
5. Explore the predictive power of you model. Provide a 2x2 cross tabulation of predictions versus the actual values and optionally display a graphic visualizing both the actual values and the predictions. Compute the total proportion of inaccurately classified individuals (= the training error) and comment on all the results. 

```{r}

# fit the model
Mod1 <- glm(high_use ~ famrel + absences + goout + sex, data = alc, family = "binomial")

# predict() the probability of high_use
probabilities <- predict(Mod1, type = "response")

# add the predicted probabilities to 'alc'
alc <- mutate(alc, probability = probabilities)

# use the probabilities to make a prediction of high_use
alc <- mutate(alc, prediction = probability > 0.5)

# see the last ten original classes, predicted probabilities, and class predictions
select(alc, famrel, absences, goout, sex, high_use, probability, prediction) %>% tail(10)

# tabulate the target variable versus the predictions
table(high_use = alc$high_use, prediction = alc$prediction) %>% prop.table %>% addmargins
```
 
