---
title: "IODS course project"
author: "Kati Puukko"
output:
  html_document:
    theme: cosmo
    toc: true
    toc_depth: 2
    fig_caption: true
    fig_width: 6
    fig_height: 4
---

***
  
```{r child = "chapter1.Rmd"}
```
  
***
  
```{r child = "chapter2.Rmd"}

```

1. Read the data from local folder and check for structure and dimensions

```{r}

library(ggplot2)
library(GGally)

students2014 <- read.table("C:/LocalData/kpuukko/IODS-project/data/learning2014.txt", sep="\t")

str(students2014)
dim(students2014)

#The data includes 7 variables and 166 observations. 
#stra, surf, stra are numerical sum variables.
#age, attitude and points are integer variables.
#gender is character variable

```
2. Show a graphical overview of the data and show summaries of the variables in the data. 

```{r}

#show graphical overview of the data
p <- ggpairs(students2014, mapping = aes(), lower = list(combo = wrap("facethist", bins = 20)))
p

#show summaries of the variables in the data

summary(students2014)

# draw a scatter plot matrix of the variables in student2014.
# [-1] excludes the first column (gender)
pairs(students2014[-1])

```
3. Choose three variables as explanatory variables and fit a regression model where exam points is the target (dependent) variable. Show a summary of the fitted model and comment and interpret the results. Explain and interpret the statistical test related to the model parameters. If an explanatory variable in your model does not have a statistically significant relationship with the target variable, remove the variable from the model and fit the model again without it. (0-4 points)

```{r}

# Linear regression model as  *stra*, *deep* and *Attitude* as predictors and *Points* as an outcome 
model1 <- lm(Points ~ stra + deep + Attitude, data = students2014)
summary(model1)

#Only attitude is statistically significant  and therefore I only include it to the second model
model2 <- lm(Points ~ Attitude, data = students2014)
summary(model2)

```
4. Using a summary of your fitted model, explain the relationship between the chosen explanatory variables and the target variable (interpret the model parameters). Explain and interpret the multiple R squared of the model. (0-3 points)

```{r}

# The parameter estimates show that attitude has positive correlation with the attitude. Participants with more positive attitude score more points.
# The multiple of R squared of the model shows that 19% of the variation is points is explained by the attitude. 

```
5. Produce the following diagnostic plots: Residuals vs Fitted values, Normal QQ-plot and Residuals vs Leverage. Explain the assumptions of the model and interpret the validity of those assumptions based on the diagnostic plots. (0-3 points)
```{r}

par(mfrow = c(2,2))
plot(model2, which = c(1, 2, 5))

# Residual vs. Fitter points spread quite randomly
#Normal Q-Q plot points fits nicely to the line 
#Residuals vs Leverage no outliers stand out 

```

***
